---
title: "Lab 1 - EDA for Cancer Dataset"
author: "Nikita Nadkarni, Thomas Drage, Venkatesh Nagapudi"
date: "September 14, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

W203 Statistics for Data Science, Fall 2018.

##1.  Introduction
This is an exploratory data analysis to examine the relationship between patient death rates caused by cancer versus various factors including the type of patient coverage, patient income levels and so on.
The data analysis includes the following sections:
1. Preliminary Dataset analysis
2. Problems with the Dataset
3. Conclusions that can be reasonably achieved
4. Recommendations 
5. Improvements to the Dataset


First include library for scatterplots
```{r}
library(car)
```


```{r load}
cancer_data = read.csv("cancer.csv")
```

### List of variables in cancer data


```{r}
(list_of_varibles = objects(cancer_data))
```
A high level summary of cancer_data shows 3047 observations of 30 variables. The important dependent variable is the deathRate. A close second is the incidence rate of cancer, which is the "avgAnnCount". 
Important indepdent variables include poverty levels of the county ("povertyPercent"), how many of them have Private or Public coverage ("PctPrivateCoverage", "PctPublicCoverage"), what is the predominant race and so on. (need to add more)

```{r}
str(cancer_data)
```





##2. Univariate Analysis of Key Variables

Thomas/Nikita - am using this section to add any problems with the dataset. We have to figure out a coherent way to structure this analysis and I am fine with any way you suggest.

a) How do we determine whether the deaths were caused because the cancer was undiagnosed? Or whether it was diagnosed and not cured because it was incurable? Or whether it was diagnosed and not cured because the patient couldn't afford it? Since everything is averaged out per county, there is no way of knowing. Perhaps in my mind, the biggest issue with this dataset is that it provides the summary data per geography, rather than the individual patient data itself. What we need is per patient data, not the summary data...perhaps we need to bring this up as the major problem. Thoughts?

Birthrate ~ births*1000 / est population

Deathrate (due to cancer) ~ deaths*100 000 / est population

b) PctPrivateCoverage and PctPublicCoverage don't add up to 100%. So some people are covered both with private and public coverage or some don't have coverage at all. The numbers range from 65 to 131 when you add them up. 
```{r}
head(cancer_data$PctPublicCoverage + cancer_data$PctPrivateCoverage, 25)
min(cancer_data$PctPublicCoverage + cancer_data$PctPrivateCoverage)
max(cancer_data$PctPublicCoverage + cancer_data$PctPrivateCoverage)
```

c) The same problem exists with the race related information. It doesn't add up to 100%. 

```{r}
head(cancer_data$PctAsian + cancer_data$PctBlack + cancer_data$PctWhite + cancer_data$PctOtherRace, 25)
min(cancer_data$PctAsian + cancer_data$PctBlack + cancer_data$PctWhite + cancer_data$PctOtherRace)
max(cancer_data$PctAsian + cancer_data$PctBlack + cancer_data$PctWhite + cancer_data$PctOtherRace)
```



##3. Analysis of Key Relationships


d) deathRate doesn't take into account the population of the county. Perhaps this is a huge problem. Maybe the data is not so bad if you find the actual % of population that died because of cancer and correlate it against other variables? A lot of the data so far that I gathered might be wrong if the main variable is not meaningful. Thoughts?

```{r}
#added new column into cancer_data
cancer_data$pctDeathRate = cancer_data$deathRate/cancer_data$popEst2015
```
Also, how do we know that the deathRate data is for 2015?

Let's see how pctDeathRate vs PcctPrivateCoverage looks now
```{r}
plot (~ jitter(cancer_data$PctPrivateCoverage,2) + jitter(cancer_data$pctDeathRate,2), type = "h")
```
How about the correlation between the two variables?
```{r}
cor(cancer_data$pctDeathRate, cancer_data$PctPrivateCoverage)
```
This shows very little correlation. (Nikita/Thomas - thoughts??)
How about the correlation between pctDeathRates and PublicCoverage?
```{r}
cor(cancer_data$pctDeathRate, cancer_data$PctPublicCoverage)
```
This seems somewhat intuitive. Perhaps with public coverage, treatment is bad.

How about median income?

```{r}
cor(cancer_data$pctDeathRate, cancer_data$medIncome)
```
A negative correlation that is as strong as the public coverage. Perhaps makes sense.

And how about poverty level

```{r}
cor(cancer_data$pctDeathRate, cancer_data$povertyPercent)
```
Not as correlated...what does this mean? That deathrates are not dependent so much on poverty?


How is death_rate correlated to the important variables? This ScatterPlotMatrix might throw some light
```{r}
scatterplotMatrix(~ deathRate + povertyPercent + medIncome + PctPrivateCoverage + PctPublicCoverage, data = cancer_data)
```
A few preliminary observations show that deathRate is positively correlated to povertyPercent and PctPublicCoverage, while it is negatively correlated to medIncome and PctPrivateCoverage. This can be verified with the correlations below:
```{r}
#correlation of deathRate to important variables
cor(cancer_data$deathRate, cancer_data$povertyPercent)
cor(cancer_data$deathRate, cancer_data$medIncome)
cor(cancer_data$deathRate, cancer_data$PctPrivateCoverage)
cor(cancer_data$deathRate, cancer_data$PctPublicCoverage)
```

There seems to be a high degree of correlation between income and deathRate. Is it possible that the deathRates are high because the cancer can be suppressable through treatment and the poorer people cannot afford it? Or is it because the cancer is incurable? We are not sure that this distinction can be reasonably figured out. But let's say that the cancer is curable - then there should be a correlation between deathRate and income. Let's see:
```{r}
plot (~ cancer_data$medIncome + cancer_data$deathRate, type = "h")
```
There seems to be a gradual drop in death rates for people with higher median incomes. However, there are definitely some outliers as we see a lot of spikes in the death rates in some cases. Are these related to incurable cancers? Are these related to the geography in some ways? How do we figure out?

Redoing some analysis here (Venky: Sep 15, 8:15pm PST) using actual numbers instead of percentages since deathRate is really not a %.

What is the correlation between deathRate and number of people with private coverage?
```{r}
NumPeopleWithPrivCoverage = cancer_data$PctPrivateCoverage * cancer_data$popEst2015
cor(cancer_data$deathRate, NumPeopleWithPrivCoverage)
```
Here is a plot 
```{r}
plot(~ jitter(log(NumPeopleWithPrivCoverage),2) + jitter(cancer_data$deathRate,2), type = "h")
```



What is the correlation between deathRate and number of people with public coverage?
```{r}
NumPeopleWithPublicCoverage = cancer_data$PctPublicCoverage * cancer_data$popEst2015
cor(cancer_data$deathRate, NumPeopleWithPublicCoverage)
```


##4. Analysis of Secondary Effects

##5. Conclusion