---
title: "Lab 1 - EDA for Cancer Dataset"
author: "Thomas Drage, Venkatesh Nagapudi"
date: "September 22, 2018"
output: 
  pdf_document: 
    fig_height: 4
    fig_width: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy=TRUE)
```

W203 Statistics for Data Science, Fall 2018.

##1.  Introduction
This is an exploratory data analysis to examine the relationship between patient death rates due to by cancer and various demographic factors including the type of insurance coverage, income level, race etc. 

These relationships will be addressed using a dataset provided by the Government Health Agency and consists of survey data taken at county level for selected areas in the United States utilising 2015 as the current population and cancer mortality data averaged across the years 2009-2013. The basis for the selection of these areas is unknown. The intent of this analysis is to assist the Government Health Agency in identifying communities which could benefit from social intervention to reduce cancer mortality rates and to identify which particular areas of intervention would be of most benefit. Note that this report does not seek to create a statistical model for the data provided but to provide insight into the key features and relationships associated with this problem.

This analysis is structured as follows:

1. Introduction

2. Univariate Analysis of Key Variables

3. Analysis of Key Relationships

4. Analysis of Secondary Effects

5. Conclusion

###Setup

This analysis is presented using the R language, utilising the following packages:
```{r}
library(car)
```

The dataset was provided in CSV format and imported for this analysis:
```{r load}
cancer_data = read.csv("cancer.csv")
```

### Data Selection

A high level summary of cancer_data shows 3047 observations of 30 variables. The important dependent variable is the deathRate. A close second is the incidence rate of cancer, which is the "avgAnnCount". 
Important indepdent variables include poverty levels of the county ("povertyPercent"), how many of them have Private or Public coverage ("PctPrivateCoverage", "PctPublicCoverage"), what is the predominant race and so on. (need to add more)
```{r}
(list_of_varibles = objects(cancer_data))
```

The datatypes can be examined and it can be noted that some level of inference is required with respect to units of measure. The data is largely numeric, with the exception of two Factor types, binnedInc which appears to be an income range for the county and Geography which is simple the state and county name.

```{r}
str(cancer_data)
```

### Data Cleansing

####a) avgAnnCount

The first variable in the dataset is the incidence rate of cancer "avgAnnCount". The outcome variable, deathRate is intuitively dependent on the incidence rate of cancer. Summarizing this variable:
```{r}
summary(cancer_data$avgAnnCount)
```
If we examine the Median and Mean, the Mean is nearly 3.5 times the median. This looks a little suspicious since we can generally expect the data distribution to be normal in nature. Here is a plot of the avgAnnCount
```{r}
plot(cancer_data$avgAnnCount, ylab="averageAnnCount")
```

There appear to be quite a few outliers present, some of which are very extreme, however, this data is not yet normalised for population. It is worthwhile to investigate the composition of this data further to see if there are any discrepancies that might be causing these outliers. From inspection of the dataset, it is clear that the avgAnnCount has some invalid data. It appears that the value "1962.667684" is repeated in several rows. We will need to discard these rows since it is unclear if the rest of the columns contain valid data. So we filter these rows out below:

```{r}
cancer_data_filter1 = cancer_data[cancer_data$avgAnnCount != 1962.667684, ]
nrow(cancer_data_filter1)
```
Now we are down to 2841 observations. 

####b) "Age" related data issues

It is safe to assume that the variable should be less than 110. If the age is greater than 110, it is likely that the data itself is invalid. We will take a look at whether Age is greater than 110 now.

```{r}
sum(cancer_data_filter1$MedianAge > 110)
sum(cancer_data_filter1$MedianAgeFemale > 110)
sum(cancer_data_filter1$MedianAgeMale > 110)
```

Clearly some of these rows could be invalid for MedianAge > 110, and it might be best to filter them out.

####c) Some issues related to average household size
We found that there are a few more invalid data items related to average household size. Clearly this has to be at least equal to or greater than 1. But we found 57 of the rows had an average household size < 0.1. We decided that these rows were untrustworthy and we filtered them out.

```{r}
sum(cancer_data_filter1$AvgHouseholdSize < 0.1)
```


```{r}
cancer_data_fixed = cancer_data_filter1[cancer_data_filter1$MedianAge < 110 &
                                        cancer_data_filter1$AvgHouseholdSize >=1, ]
nrow(cancer_data_fixed)
```

Now we are down to 2755 observerations.

####c) Potential problems with Private and Public Coverage not adding up to 100%
PctPrivateCoverage and PctPublicCoverage don't sum to 100% in the majority of cases. This implies that people are covered both with private and public coverage or some don't have coverage at all. 

The sum of both groups ranges from 65 to 131 which is not unreasonable and while this is likely not a problem, we do want to point out this that there could be a possible discrepancy, especially if the incidence of having both forms of coverage was expected to be low.

```{r}
min(cancer_data$PctPublicCoverage + cancer_data$PctPrivateCoverage)
max(cancer_data$PctPublicCoverage + cancer_data$PctPrivateCoverage)
```

####d) Problem with incomplete race related information
The same problem exists with the race related information. It doesn't sum to 100% for many counties, however it is always less than 100, and so it appears some people elected not to state their race. Mostly Alaskan counties form the outliers - the predominant race is likely not captured by this survey.

```{r}
head(cancer_data$PctAsian + cancer_data$PctBlack + cancer_data$PctWhite +
       cancer_data$PctOtherRace, 25)
min(cancer_data$PctAsian + cancer_data$PctBlack + cancer_data$PctWhite +
      cancer_data$PctOtherRace)
max(cancer_data$PctAsian + cancer_data$PctBlack + cancer_data$PctWhite +
      cancer_data$PctOtherRace)
```

####e) Source of marriage data not clear

The dataset gives family information in two formats, PercentMarried and PctMarriedHouseholds. Whilst the exact definitions are not available, we can assume that the the former refers to invididuals and the latter counted if a married couple exists withint he household. Notwithstanding factors such as multiple occupancy and separations we would expect these to be highly correlated. The summaries below support a reasonable range and distribution for this data as-is.

```{r}
summary(cancer_data$PercentMarried)
summary(cancer_data$PctMarriedHouseholds)
```

####f) Education data non-hierarchical and categorisation mixed

Education demographic data is given in a series of columns, categorised by education level and age of respondant. This appears to be raw (but aggregated) survey data. The available columns are:

* PctNoHS18_24: High school not completed, ages 18-24.
* PctHS18_24: High school completed, ages 18-24.
* PctHS25_Over: High school completed, over 25.
* PctSomeCol18_24: Some college completed, ages 18-24.
* PctBachDeg18_24: Bachelor's degree obtained, ages 18-24.
* PctBachDeg25_Over: Bachelor's degree obtained, over 25.

* PctEmployed16_Over: Employment rate, over 16.
* PctUnemployed16_Over: Unemployment rate, over 16.

Note that due to these definitions it is difficult to verify completeness as there are groups we have no information for, e.g. high school not completed for over 25s or those with some college over 25. It should be noted that the strict definitions do not represent vocational education or higher degrees either. We can verify the distribution is reasonable however;

```{r}
summary(cancer_data$PctNoHS18_24)
summary(cancer_data$PctHS18_24)
summary(cancer_data$PctHS25_Over)
summary(cancer_data$PctSomeCol18_24)
summary(cancer_data$PctBachDeg18_24)
summary(cancer_data$PctBachDeg25_Over)
```

The significant number of NA records for PctSomeCol18_24 renders this field unusable for analysis. The zero minimums for PctHS18_24 and PctBackDeg18_24 are also questionable, the single record showing no high school graduates should be excluded from individual analysis however the 118 counties with no respondants without a degree are problematic and will need to be considered when drawing conclusions from this type of data.

```{r}
sum(cancer_data$PctHS18_24 < 0.01)
sum(cancer_data$PctBachDeg18_24 < 0.01)
```

The categorisation of unemployment data also leaves something to be desired as the rest of the population either unable to work or retired is not represented, however on a basic level the data seems reasonable.

```{r}
summary(cancer_data$PctEmployed16_Over)
summary(cancer_data$PctUnemployed16_Over)
```

##2. Univariate Analysis of Key Variables

###a) Establishing an incidence rate using avgAnnCount

The "avgAnnCount" variable is not insightful as given, as it needs to be averaged over the population of the county. For comparison purposes, it needs to have the same defintion of the key outcome variable, deathRate. 

If we look at the above summary, for the first row, BirthRate is 6.12, while deathRate is 165. The avgAnnCount is 1397. 

In order to get the definition of the units of measure for each of these variables, as they were not provided in the dataset we have utilised and verified against the typical ways of representing these measures:

BirthRate is the number of births per 1000 people,
DeathRate is the number of deaths per 100,000 people,
avgAnnCount is the raw number of cancer incidents in the county.

To prepare avgAnnCount, we have used the same definition as deathRate and created a new variable, "incidenceRate":
```{r}
incidenceRate = 100000*cancer_data_fixed$avgAnnCount/cancer_data_fixed$popEst2015
summary(incidenceRate)
```

The mean and median values are very close indicating a symmetric distribution, which can be verified in a histogram:

```{r}
hist(incidenceRate, main = "Incidence Rate of Cancer", 
     xlab = "Incidence Rate per 100,000 people")
```

It is worth noting that in this format the significant outliers noted in averageAnnCount have been eliminated due to the population adjustment. The standard deviation of this quantity is: 

```{r}
sd(incidenceRate)
```
Which shows a significant variation in incidence on a geographical basis, foreshadowing some demographic dependence on cancer incidence in the US.

###b) Analyzing the key outcome variable: deathRate

Now looking at the key outcome variable deathRate and summarizing it:
```{r}
summary(cancer_data_fixed$deathRate)
```
The Median and Mean are very close indicating a symmetric distribution. Plotting this variable in a histogram:
```{r}
hist(cancer_data_fixed$deathRate, main = "Death Rate due to cancer", 
     xlab = "Death rate per 100,000 people")
```

This shows that on an average 180 people die of cancer while 550 people are diagnosed with cancer for every 100,000 people. This means that nearly 1/3rd of all people who are diagnosed with cancer die within six years of diagnosis. 

The standard deviation is:

```{r}
sd(cancer_data_fixed$deathRate)
```
This is slightly narrower on a percentage basis (16% vs 22%) than the cancer incidence rate and the relationship between these two quantities is worthy of further analysis in Section 3. 

(Note: The study indicates that the deathRates are from 2015 while the incidence rates are between 2009 and 2013.) 

###c) Distribution of birthRate and median age

For completeness, the birthRate is analysed here, though it seems very unlikely that this will influence the deathRate due to cancer. Summarizing the birthRate:
```{r}
summary(cancer_data_fixed$BirthRate)
```
The mean and median are similar, though the maximum seems to be very high. Plotting the data with a histogram:
```{r}
hist(cancer_data_fixed$BirthRate, main = "Birth Rate Distribution", breaks = 0:22 - 0.5,
     xlab = "Birth Rate in for 1000 people")
```
Birth rates over 12 seem to be very low in terms of frequency. 

The average birth rate of approximately 5.0 is is questionable since the average birth rate in the USA is around 12.2 per 1,000 people. It is unclear why these counties have a low birth rate that is more than 60% lower than national average. Perhaps there is a certain type of cancer that causes lower birth rates? This anomaly would benefit from further investigation.

There is a reasonably useful age spread in the median age of counties, which will allow us to investigate effects of age on cancer incidence. The distribution peaks around middle age as expected and there are few communities with medians at the age-extremes.

```{r}
hist(cancer_data_fixed$MedianAge, main = "County Median Age Distribution", 
     xlab = "Median Age")
```

###d) Analyzing how patients are covered by Private and Public Coverage

As mentioned earlier, it seems as if some patients have both Public and Private Coverage, or some have neither. On average however, slightly more than 60% of people have public insurance and slightly less than 40% have private insurance and the degree of variation between counties is not particularly large. Some counties have significant variation, potentially related to emplyoment or economic factors.

```{r}
boxplot(cancer_data_fixed$PctPublicCoverage, cancer_data_fixed$PctPrivateCoverage, 
        main = "Public and Private Insurance Coverage Distribution", 
        names = c("Private","Public"), ylab = "Percent")
```
\newline
Analyzing Private Coverage further, it is clear that most private coverage is employer provided.
```{r}
hist(cancer_data_fixed$PctEmpPrivCoverage / cancer_data_fixed$PctPrivateCoverage,
     main = "Percentage of Employer Provided Private Coverage", 
     xlab = "Percentage Employer provided Private coverage")
```

###e) Analyzing Race related information

Further analysis of the race-related information shows that it is definitely incomplete, for example, we can examine the distribution of races and the distribution of the sum of the given categories;

```{r}
boxplot(cancer_data$PctWhite, cancer_data$PctBlack, cancer_data$PctAsian,
        cancer_data$PctOtherRace, cancer_data$PctWhite + cancer_data$PctBlack +
        cancer_data$PctAsian + cancer_data$PctOtherRace, 
        names = c("White", "Black", "Asian", "Other", "Sum"))
```

There is definitely a predominance of caucasians in the population as expected, but the inconsistency in the sum points to a preference not to specify race in the survey. As it is well known that genetic predispositions to diseases linked to race exist this is problematic for our analysis as some minorities are certainly under-represented. 

###f) Analyzing and preparing Income related variables

Let's analyze medIncome and binnedInc. medIncome is the median income of the county is a good indicator on the poverty percent. Binned Income is the income separated out into 10 bins. 

Summarizing the two variables, here is what we see
```{r}
summary(cancer_data_fixed$medIncome)
summary(cancer_data_fixed$binnedInc)
```

The binned income has a problem as the last level is not ordered. We will need to put the last bin first. So we can prepare the data as follows:
```{r}
cancer_data_fixed$binnedInc = factor(cancer_data_fixed$binnedInc, 
                                     c("[22640, 34218.1]","(34218.1, 37413.8]",
                                       "(37413.8, 40362.7]","(40362.7, 42724.4]",
                                       "(42724.4, 45201]","(45201, 48021.6]",
                                       "(48021.6, 51046.4]","(51046.4, 54545.6]",
                                       "(54545.6, 61494.5]","(61494.5, 125635]"))
```


##3. Analysis of Key Relationships

###a) Relationship between incidence rate and death rate
We now examine how the key outcome variable deathRate is related to other variables. First, let's see if there is a correlation between the incidence rate and death rates. Intuitively, it seems to make sense that the higher the incidence rate, the higher the death rate should be, however we noted a variation in the distributions previously.  

```{r}
plot(incidenceRate, cancer_data_fixed$deathRate)
```

There is a general trend that higher death rate due to cancer occurs where there is a higher incidence rate, but the significant variability observed indicates that there are other effects at play which are independent of the two events. For example, environmental or race-prevalance effects may influence cancer incidence but health insurance or poverty effects may have a greater role in death due to variability of treatment outcomes. We can quantify this with the correlation coefficient:

```{r}
cor(incidenceRate, cancer_data_fixed$deathRate)
```

This indicates a positive correlation, however it is not particularly strong.

###b) Relationship between death rate and median income

One of the key relationships we want to examine is whether a county with a lower income generally sees a higher deathRate since it is likely that battling cancer is an expensive proposition. First we examine the correlation:
```{r}
cor(cancer_data_fixed$deathRate, cancer_data_fixed$medIncome)
```
There is a negative correlation confirming that the deathRate is indeed higher if the median income is lower. 

Let's now figure the correlation between deathRate and median income bins using a boxplot.

```{r}
boxplot(deathRate ~ binnedInc, data = cancer_data_fixed, 
        main = "Death rate vs Median Income bins",
        xlab = "Median Income bins",
        ylab = "Death Rate")
```
As can be clearly seen, the higher the income, the lower is the death rate. This indicates that if the county is very poor on an average, then it is likely to have a higher death rate. We test this hypothesis by comparing deathRate against povertyPercent. 

Revisting (a) above, we can note the correlation with incidenceRate:

```{r}
cor(incidenceRate, cancer_data_fixed$medIncome)
```
This is a lesser correlation and highlights that a larger variety of factors are at play in causing incidences of cancer than in treatment outcomes.

```{r}
cor(cancer_data_fixed$povertyPercent, cancer_data_fixed$deathRate)
```
We notice that the correlation to povertyPercent is almost identical in absolute value to the median income (medIncome). 
Let's examine the correlation between medIncome and povertyPercent now:
```{r}
cor(cancer_data_fixed$medIncome, cancer_data_fixed$povertyPercent)
```
There is a very strong negative correlation nearly close to -1.0 as expected between medIncome and povertyPercent. This similarity suggests only that they both indicate the same economic influence on cancer death rate.

###c) Relationship between death rate and private & public coverage

Clearly the affordability of healthcare is a major consideration is the battle against cancer. It is likely that if a person has a higher salary, he will choose to go with Private Health care which is generally regarded to provide better services than Public Health care. we can examine the correlation between these two variables.

```{r}
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctPrivateCoverage)
```
This does show the negative correlation confirming that deathRates are inversely related to Private Coverage. This can be verified in a plot:

```{r}
plot (~ PctPrivateCoverage + deathRate, data = cancer_data_fixed, 
      col="red", xlab="% Coverage")
abline(lm(cancer_data_fixed$deathRate ~ cancer_data_fixed$PctPrivateCoverage), col="red")
points (cancer_data_fixed$PctPublicCoverage, cancer_data_fixed$deathRate, col="blue")
abline(lm(cancer_data_fixed$deathRate ~ cancer_data_fixed$PctPublicCoverage), col="blue")
legend("topright", legend = c("Private", "Public"), col = c("red", "blue"), pch = c(19,19))
```


It is evident above, however, that the relationship with public coverage is quite different and we can now examine the relationship between deathRate and Public coverage:
```{r}
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctPublicCoverage)
```
There is a very strong positive correlation indicating that death rates in counties that have higher Public Coverage are greater. This is possibly due to incidence of public health coverage indicating lower subscription to private health insurance, but it would be expected to be tempered by those who have no insurance. It appears in any case that it is a problem if the county relies on public health coverage.

To quantify this we can examine the gain associated with private health insurance uptake:

```{r}
lm(100*cancer_data_fixed$deathRate/mean(cancer_data_fixed$deathRate) 
   ~ cancer_data_fixed$PctPrivateCoverage)
```
This indicates that, for example, a 10% greater uptake in private health insurance is associated with a 5.5% reduction in the county's death rate due to cancer.


###d) Importance of age on incidence rate and death rate

Does cancer affect older people much more than younger ones? This might be important so that we can understand whether younger people need to have some sort of coverage so that they don't succumb to cancer.

For this, let's examine incidenceRate of cancer against age.

```{r}
cor(incidenceRate, cancer_data_fixed$MedianAge)
```
There is a very large correlation between incidence rate and median age. It seems that higher the age, the greater there is chance of getting cancer. We can quantify this in a useful sense:

```{r}
lm(incidenceRate ~ cancer_data_fixed$MedianAge)
```

This reveals that, a county with median age around 50 typically saw twice the cancer cases as one with a median age around 30. The same treatment can be repeated for the deathRate:

```{r}
cor(cancer_data_fixed$deathRate, cancer_data_fixed$MedianAge)
```

There is almost no correlation between the two variables! Clearly, while a county with greater median agetends to have a greater cancer incidence, age itself is not a significant factor when it comes to the deathRate, at least in the insight gained from county level aggregate statistics without the ability to correlate patient outcomes within age brackets.

But it might make sense to examine whether age affects the likelihood of getting private coverage. Perhaps if it is easy for the younger people to get private coverage, that might explain why MedianAge does not correlate well with deathRate. Plotting the two variables out:

```{r}
plot(~ cancer_data_fixed$MedianAge+cancer_data_fixed$PctPrivateCoverage, 
     xlab = "Age", 
     ylab = "Pct Private Coverage" )
```

Clearly, there is no linear relationship present above, indicating that age is not a factor in getting private coverage. However, we note that there is a demographic split present at counties with younger median age, strong clustering around middle age (likely when employer provided insurance is available) and then a less varied spread in the older communities.


##4. Analysis of Secondary Effects

###a) The importance of race on incidence rate and death rate

One of the interesting points to understand is whether a certain demographic is poorer than others, and if so, whether that is causing a significant impact to the deathRate. This could result in actionable insights for certain organizations (say charities) that are focused on a particular race. For this purpose, we can examine the correlation between PovertyPercent and the race.
```{r}
cor(cancer_data_fixed$povertyPercent, cancer_data_fixed$PctBlack)
cor(cancer_data_fixed$povertyPercent, cancer_data_fixed$PctWhite)
cor(cancer_data_fixed$povertyPercent, cancer_data_fixed$PctAsian)
cor(cancer_data_fixed$povertyPercent, cancer_data_fixed$PctOtherRace)
```
The results reveal that the presence of the black race group is positively correlated with higher cancer incidence whilst the white race group is negatively correlated and also, though to a lesser extent the presence of asian race group. This can be compared to the correlation with death rate due to cancer:

```{r}
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctBlack)
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctWhite)
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctAsian)
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctOtherRace)
```
This reveals a similar inversion in the correlation for the black race group comapred to others. An important result here is that the government could focus on improving the conditions or, for example, investigating genetic predispositions of members of this group and hence reduce cancer death rates in these counties.

We can repeat such an analysis for the incidence of cancer:

```{r}
cor(incidenceRate, cancer_data_fixed$PctBlack)
cor(incidenceRate, cancer_data_fixed$PctWhite)
cor(incidenceRate, cancer_data_fixed$PctAsian)
cor(incidenceRate, cancer_data_fixed$PctOtherRace)
```

This is a very interesting insight and shows that presence of caucasians is positively associated with cancer incidence. This result supports the hypothesis of poverty differentiating access to quality healthcare, but also highlights that difference race groups may be affected by different types of cancer with a specific range of outcomes most typical for each.


###b) Factors affecting choice of insurance coverage

Since private health care coverage appears to have a positive impact on cancer outcomes, it is interesting to investigate the factors driving choice - do people choose private coverage if they are above the poverty level or with some education level? And do they choose to go with public coverage otherwise? We can examine this below:
```{r}
scatterplotMatrix(~ cancer_data_fixed$povertyPercent + 
                    cancer_data_fixed$PctPrivateCoverage +
                    cancer_data_fixed$PctPublicCoverage + 
                    cancer_data_fixed$PctHS25_Over + 
                    cancer_data_fixed$PctBachDeg25_Over)
```
\newline
It is evident that people who are below the poverty line or less educated (note the two are interrelated) typically utilise public insurance coverage. In general, this supports the view that economic factors, whether by preventative means or by access to improve treatment via private healthcare lead to better cancer outcomes.

We can follow this by examining education directly in relation to death rate due to cancer. For this purpose we select two variables - percent of people with Bachelor's degrees and people with just High School education (over 25) and observe the correlation to Death rate. Here the assumption is that these represent the highest educational degrees completed by the people. For the purposes of exploratory data analysis these are good indicators of the education level in the age groups.

```{r}
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctBachDeg25_Over)
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctHS25_Over)
```

There is definitely a significant positive correlation between death rate and highest education being HighSchool. And if the education is higher, it seems like deathRate falls.
This is likely because higher education would lead to better employment, which could result in private coverage, and better means to fight cancer. Let's examine the relationship between employment and education to be sure this is the case.

```{r}
cor(cancer_data_fixed$PctEmployed16_Over, cancer_data_fixed$PctBachDeg25_Over, use = "complete.obs")
cor(cancer_data_fixed$PctEmployed16_Over, cancer_data_fixed$PctHS25_Over, use = "complete.obs")
```
In this case, we had to use "complete.obs" since there were some invalid observations. As we can clearly see, having a Bachelor's degree seems to be positively correlated to being employed, supporting our chain of reasoning.

###d) Does being married help with reduction of deathRates?

What one can intuit is that married people might have better coverage than unmarried ones, and that could lead to reduced deathRates. Let's examine this relationship. Firstly, what is the correlation between marriage and coverage?

```{r}
cor(cancer_data_fixed$PctMarriedHouseholds, cancer_data_fixed$PctPrivateCoverage)
cor(cancer_data_fixed$PctMarriedHouseholds, cancer_data_fixed$PctPublicCoverage)
```
The data seems to make sense. It is likely that the median income is higher in the case of married families. Let's see if that is the case with a boxplot

```{r}
boxplot(PctMarriedHouseholds ~ binnedInc, data = cancer_data_fixed, 
        main = "Married Households vs Median Income bins",
        xlab = "Median Income bins",
        ylab = "Married Households")
```
\newline
We note that the expected correlation between presence of married households and income exists.

So that we complete this picture, if we compare deathRates against MarriedHouseholds, there should be a negative correlation:
```{r}
cor(cancer_data_fixed$deathRate, cancer_data_fixed$PctMarriedHouseholds)
```

It is indeed so; DeathRates are lower when patients are married, as seen above, most probably due to higher median incomes allowing better treatment outcomes or improved lifestyle/preventative measures taken by married couples. Further research would be required to establish the specific causality.

###d) Does size of household have a bearing on deathRates?
Let's examine the size of households now.
```{r}
summary(cancer_data_fixed$AvgHouseholdSize)
```

Average household size ranges from 1.86 through 4 people per household after filtering for extremely low values which were implausible as an average of integers.

Let's plot deathRate against the number of people in a boxplot to see the correlation.

```{r}
HouseholdSize_bin = cut(cancer_data_fixed$AvgHouseholdSize, 
                        breaks = c(1, 2, 3, 4, Inf))
boxplot( deathRate ~ HouseholdSize_bin , data = cancer_data_fixed,
        main = "Death rate vs Average Household size",
        xlab = "Average Household Size",
        ylab = "Death Rate")
```

In communities with greater average household size, we observe a lesser the death rate. It could be that the median income is higher in the case of the bigger households and we can investigate:

```{r}
boxplot(AvgHouseholdSize ~ binnedInc, data = cancer_data_fixed, 
        main = "Average household size vs Median Income bins",
        xlab = "Median Income bins",
        ylab = "Average household size")
```

There is an increase in median income as the average household size increases, but only slightly so and it is likely that the causality here is due to greater income enabling the economics of having a larger family. The correlation with death rate may instead be explained due to the typical presence of married couples in larger households and the presence of children in large households, and per our earlier investigations, a shift towards younger age groups is expected to result in lower incidence of cancer.


##5. Conclusion